Design and implemention of open health data systems with standardized data and analytics capabilities for low- and middle income countries

### 2024-08-27 index.qmd
Add table 1 with types of data sharing

Digital transformation should be considered as a key determinant of health, but also requires a radical rethink how digital technologies should be used [@kickbusch2021lancet]. Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services as we know them; change nature and direction of health care, and the possibility of making it more preventive, personalised, and mobile; and enfranchise patients and communities (particularly vulnerable groups), thus helping shape the health system according to the needs of the patients and communities.

@@

Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions is within reach. As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs. By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of maternal, newborn and child health (MNCH) services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].

MomCare, like most other digital health interventions, is critically dependent on the availability of data and analytics functions within its underlying supportive OHDS. For example, the widely used DHIS2 reports that are most commonly used in LMICs contained insufficient information and analytical functions to support such interventions. To implement the value-based healthcare business logic, more detailed analytics of patient journeys is required.

It is in fact a widely held belief in the global digital health community that data & analytics services are an essential capibility of OHDSs. This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one]. These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper]. In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond it original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies etc. If we are to use the OpenHIE framework for data & analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.

@@

- **Ingestion Services:** use of FHIR standard to harmonize all incoming healthcare data to a common data model, including metadata extraction and metadata modeling. Should support both single-records streaming ingest as well as bulk data ingestion in batches using the Bulk FHIR API as interface [@mandl2020push]. 
- **Storage Service:** should support columnar storage engines optimized for analytical workloads. In case of file-based storage, use columnar file formats such as Apache Parquet. In case of databases, prefer open source engines such as Clickhouse or PostgreSQL that also support external, file-based tabels [@pedreira2023composable].
- **Maintenance:** use one of the open table formats such as Apache Iceberg, Apache Hudi and/or Delta Lake [@jain2023analyzing] to realize dataset organization, data integration, schema evolution and data provenance. Use of SQL-on-FHIR v2 View definitions to facilitate access to FHIR resources in flattened, tabular format [@sql-on-fhir]
- **Exploration:** use on-demand, read-only analytical processing engines to provide a unified querying interface to access the heterogeneously structured data using new dataprocessing technologies such as DuckDB, polars etc. Should support SQL-on-FHIR Runners, such as Pathling, to generate the standardized views on demand.

>>>>

It is a widely held belief in the global health community that digital technologies have an important role to play in strengthening healthcare in low- and middle income countries (LMICs). Digital technologies have the potential to increase the availability, accessibility, acceptability, and quality of health services; make healthcare more preventive, personalised, and mobile; and enfranchise patients and communities, particularly those who are most vulnerable [@kickbusch2021lancet].

@@

Learnings from digital health interventions have shown that improvements along the five aforementioned dimensions are within reach. As an example, in our experience with the MomCare programme, we have demonstrated that routinely collected health data, combined with financial data, can be effectively used to gain insight into the continuity of care, improve clinical adherence whilst maintaining efficiency and affordability of health services in LMICs. By actively coaxing pregnant mothers to undergo 4 antenatal check-ups, outcomes were improved whilst maintaining the average cost of maternal, newborn and child health (MNCH) services [@huisman2022digital;@sanctis2022maintaining;@izudi2023experiences].

MomCare is critically dependent on the availability of data and analytics functions within its underlying supportive OHDS. For example, to implement the value-based healthcare business logic of MomCare, detailed analysis of patient journeys is required. This functionality is currently not available as a standard; standardized reports that are available, such as DHIS2, contained insufficient information and analytical functions to support intervention that aim to improve continuity of care and adherence to clinical guidelines. As such, a large part of the day-to-day operation of MomCare revolved performaing data and analytics: data acquisition, data integration, analysis etc.

This situation is not unique to MomCare. In fact, many in the global digital health community think that data & analytics services should be an essential capibility of OHDSs going forward. This is exemplified how the OpenHIE reference architecture [@openhie] has been adopted by many sub-Saharan African countries as the blueprint for implementing nation-wide health information exchanges (HIE) [@mamuye2022health], including Nigeria [@dalhatu2023paper], Kenya [@mbugua2021adoption] and Tanzania [@nsaghurwe2021one]. These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain [@mbugua2021adoption;@dalhatu2023paper]. In terms of the often-used distinction between primary and secondary health data use [@cascini2024health], these countries aim to extend OpenHIE beyond it original scope of primary data sharing to also include secondary use of health data for academic research, real-world evidence studies etc. If we are to use the OpenHIE framework for secondary use, supported by data & analytics as well, we need to extend the standards, technologies and architecture to include functionality to do so. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish more comprehensive platforms to support primary and secondary health data sharing in LMICs.

### Objects of openness in secondary data sharing

To set the scene of for the main contributions of this viewpoint paper, consider four types of secondary data sharing as shown in @tbl-types-data-sharing, where we follow the research agenda proposed by de Reuver et al. to scrutinize how openness of data platforms can be achieved [@dereuver2022openness].

|   | __Type of data sharing__ | __Relevance to extension of OpenHIE specification__ |
|:-:|:-------|:-------------|
| __1__ | Data at the most granular level with which the patient journey (timeline) can be reconstructed and used for various analytic tasks. | The Shared Health Record (SHR) is specified as an operational, real-time transactional data source, distinct from a data warehouse. A seperate specification of data and analytics functions, typically provided by a datawarehouse, is required. |
| __2__ | Aggregated data, typically used for routine reports and benchmarking. | The Health Management Information System (HMIS) should support Aggregate Data Exchange (ADX) workflow standard. More flexible and extensive workflows are emerging based on FHIR. | __SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows. Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items. Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
| __3__ | Data analytics modules, that provide secure and privacy-preserving computational environments to work with the data. | Federated learning (FL) [@rieke2020future] and privacy- enhancing technologies (PETs) [@scheibner2021revolutionizing;@jordan2022selecting] provide new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself. Requires use of a common data model, such as FHIR, to analyze the data in a collaborative, decentralized fashion. |
| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | Increasing need to open source trained AI models [@osi2024open], enabled by technologies such as ONNX [@onnx]. |

: Types of secondary data sharing, and there relevance to the proposed extension of the OpenHIE specification. {#tbl-types-data-sharing .striped .hover .responsive}

[TO DO: add more text here to clarify our position and intention with this paper.]

### Outline

@@

- **Ingestion Services:** use of FHIR standard to harmonize all incoming healthcare data to a common data model, including metadata extraction and metadata modeling. Should support both single-records streaming ingest as well as bulk data ingestion in batches using the Bulk FHIR API as interface [@mandl2020push;@jones2021landscape]. 
- **Storage Service:** should support columnar storage engines optimized for analytical workloads. In case of file-based storage, use columnar file formats such as Apache Parquet. In case of databases, prefer open source engines such as Clickhouse or PostgreSQL that also support external, file-based tabels [@pedreira2023composable].
- **Maintenance:** use one of the open table formats such as Apache Iceberg, Apache Hudi and/or Delta Lake [@jain2023analyzing] to realize dataset organization, data integration, schema evolution and data provenance. Use of SQL-on-FHIR v2 View definitions to facilitate access to FHIR resources in flattened, tabular format [@sql-on-fhir]
- **Exploration:** use on-demand, read-only analytical processing engines to provide a unified querying interface to access the heterogeneously structured data using new dataprocessing technologies such as DuckDB, polars etc. Should support SQL-on-FHIR Runners, such as Pathling, to generate the standardized views on demand.
